### 2022-02-23 ehoh/ehoh.tex
index

@@

>>>>

\subsection{Matching}

Given $s$ and $t$, the matching problem consists of finding a
substitution $\sigma$ such that $\sigma(s) = t.$
We then write that ``$t$ is an instance of $s$'' or
``$s$~generalizes~$t$.'' %and call $\sigma$ a \emph{generalizing substitution}.
We are interested in most general generalizations (MGGs).
Matching can be reduced to unification by treating variables in $t$ as
nullary symbols \cite{bn-98-tr-and-all-that}, but E implements it
separately.

Matching can be specified abstractly as a transition system on matching constraints
$s_i \MATCH t_i$ consisting of the
unification rules \textsf{Decompose},
\textsf{DecomposeX},
\textsf{Clash},
\textsf{ClashTypeX},
\textsf{ClashLenXF}
(with $\MATCH$ instead of $\unif$)
and augmented with

\noindent
\begin{tabular}{ll}
  \textsf{Double} &   $\{ x \MATCH t{,}\; x \MATCH t' \} \uplus \mathit{S} \Matcharrow \bot$; if $t \not= t'$  \\[\jot]
  \textsf{ClashLenXY} & $\{ x \; \tuple{s}{m} \MATCH y \; \tuple{t}{n} \} \allowbreak \uplus \mathit{S} \Matcharrow \bot$; if $x \not= y$ and $m > n$ \\[\jot]
  \textsf{ClashFX} & $\{ \cst{f} \; \overline{s} \MATCH x \; \overline{t} \} \uplus \mathit{S} \Matcharrow \bot$
\end{tabular}

The matching relation is sound, complete, and well founded. Interestingly, a
\textsf{Delete} rule would be unsound for matching. Consider the problem
$\{ x \MATCH x, x \MATCH \cst{g} \; x \}$. Applying \textsf{Delete} to the
first constraint would yield the solution $\{ x \MATCH \cst{g} \; x \}$, even
though the original problem is clearly unsolvable.

@@

Superposition provers like E work by saturation. Their main loop
heuristically selects a clause and searches for potential inference
partners among a possibly large set of other clauses. Mechanisms such
as simplification and subsumption also require locating terms in a
large clause set. For example, when E derives a new equation
$s \eq t$, if $s$ is larger than $t$ according to the term order, it
will rewrite all instances $\sigma(s)$ of $s$ to $\sigma(t)$ in
existing clauses.

To avoid iterating over all terms (including subterms) in large clause
sets, superposition provers store the potential inference partners in
indexing data structures. A term index stores a set of terms $\mathit{S}.$
Given a \emph{query term}
%
$t$, a query returns all terms $s \in \mathit{S}$ that satisfy a given
\emph{retrieval condition}:\ $\sigma(s) = \sigma(t)$ ($s$~and $t$ are
unifiable), $\sigma(s) = t$ ($s$ generalizes~$t$), or $s = \sigma(t)$
($s$ is an instance of $t$), for some substitution~$\sigma.$
\emph{Perfect} indices return exactly the subset of terms satisfying
the retrieval condition. In contrast, \emph{imperfect} indices return
a superset of eligible terms, and the retrieval condition needs to be
checked for each candidate.

E relies on two term indexing data structures, perfect discrimination
trees \cite{mcc-92-pdts} and fingerprint indices
\cite{ss-12-fp-indexing}, that needed to be generalized to
\lfhol. It also uses feature vector indices
\cite{ss-2013-feature-vector} to speed up subsumption and
related techniques, but these require no changes to work
with \lfhol{}.

\subsection{Discrimination Trees}

\looseness=-1
Discrimination trees \cite{mcc-92-pdts} are tries in which every
node is labeled with a symbol or a variable. A path from the root to
a leaf corresponds to a ``serialized term''---a term expressed without
parentheses and commas. Consider the following discrimination trees $D_1$ and
$D_2$:
%
\begin{align*}
 & \vcenter{\hbox{\begin{tikzpicture}[level distance=2.25em, sibling distance=5em,
%  edge from parent path={(\tikzparentnode.south) -- ++(0,-0.5em)
%   -| (\tikzchildnode.north)},
  every node/.style = {shape=circle, minimum size=1.75em, draw, align=center}]
  \node {\phantom{x}}
    child { node {$\cst{f}$}
      child { node %[fill=verylightgray]
        {$\cst{a}$} }
    }
    child { node % [fill=verylightgray]
      {$\cst{g}$}
      child { node {$\cst{a}$}
        child { node % [fill=verylightgray]
          {$\cst{a}$} }
      }
      child[sibling distance=2.5em] { node {$\cst{b}$}
        child { node %[fill=verylightgray]
          {$\cst{a}$} }
        child { node %[fill=verylightgray]
          {$\cst{b}$} }
      }
    }
    ;
\end{tikzpicture}
}}
&&
  \vcenter{\hbox{\begin{tikzpicture}[level distance=2.25em, sibling distance=5em,
%  edge from parent path={(\tikzparentnode.south) -- ++(0,-0.5em)
%   -| (\tikzchildnode.north)},
  every node/.style = {shape=circle,
    draw, align=center}]
  \node {\phantom{x}}
    child { node {$\cst{f}$}
      child { node %[fill=verylightgray]
        {$x$} }
    }
    child { node % [fill=verylightgray]
      {$\cst{g}$}
      child { node {$\cst{a}$}
        child { node % [fill=verylightgray]
          {$\cst{a}$} }
      }
      child[sibling distance=2.5em] { node {$y$}
        child { node %[fill=verylightgray]
          {$\cst{a}$} }
        child { node %[fill=verylightgray]
          {$x$} }
      }
    }
    child[sibling distance=3.75em] { node {$x$} }
    ;
\end{tikzpicture}
}}
\end{align*}
%
Assuming $\cst{a}, \cst{b}, x, y : \iota$, $\cst{f} : \iota \to \iota$,
and $\cst{g} : \iota^2 \to \iota$,
$D_1$ represents the term set
$\{
\cst{f}(\cst{a})
{,}\allowbreak\;
\cst{g}(\cst{a}, \cst{a})
{,}\allowbreak\;
\cst{g}(\cst{b}, \cst{a})
{,}\allowbreak\;
\cst{g}(\cst{b}, \cst{b})
\}$,
and $D_2$ represents the term set
$\{
\cst{f}(x)
{,}\allowbreak\;
\cst{g}(\cst{a}, \cst{a})
{,}\allowbreak\;
\cst{g}(y, \cst{a})
{,}\allowbreak\;
\cst{g}(y, x)
{,}\;
x
\}$.
%
E uses perfect discrimination trees for finding generalizations of
query terms. Thus, if the query term is
$\cst{g}(\cst{a}, \cst{a})$, it would follow the path $\cst{g}.\cst{a}.\cst{a}$
in~$D_1$ and return $\{\cst{g}(\cst{a}, \cst{a})\}.$
For~$D_2$, it would also explore paths labeled with
variables, binding them as it proceeds, and return $\{
\cst{g}(\cst{a}, \cst{a}){,}\allowbreak\; \cst{g}(y, \cst{a}){,}\allowbreak\; \cst{g}(y, x){,}\; x \}.$

It is crucial for this data structure that distinct terms always give rise to
distinct serialized terms. Conveniently, this property also holds for \lfhol{}
terms. Suppose that two distinct \lfhol{} terms yield the same serialization.
Clearly, they must disagree on parentheses; one will have the subterm $s\; t\;
u$ where the other has $s\; (t\; u).$ However, these two subterms cannot both
be well typed.

\looseness=-1
When generalizing the data structure to \lfhol, we face a complication
due to partial application. First-order terms can only be stored in leaf nodes,
but in Ehoh we must also be able to represent partially applied terms,
such as $\cst{f}$, $\cst{g}$, or $\cst{g}\;\cst{a}$ (assuming, as above, that
$\cst{f}$ is unary and $\cst{g}$ is binary). Conceptually, this can be solved
by storing a Boolean on each node indicating whether it is an accepting state.
In the implementation, the change is more subtle, because several parts of E's
code implicitly assume that only leaf nodes are accepting.

\newcommand\TermStack{\ensuremath{T}}
\newcommand\BTStack{\ensuremath{P}}

The main difficulty specific to \lfhol{} concerns applied variables. To
enumerate all generalizing terms, E needs to backtrack from child to
parent nodes. This is achieved using two stacks that store subterms of
the query term:
\TermStack{} %(\texttt{term\_\allowbreak stack})
stores the terms that
must be matched in turn against the
current subtree, and
\BTStack{} %(\texttt{term\_\allowbreak proc})
stores, for each node from
the root to the current subtree, the corresponding processed term.

Let $[a_1, \dots, a_n]$ denote an $n$-item stack with $a_1$ on top. Given a
query term $t$, the matching procedure starts at the root with $\sigma =
\emptyset$, $\TermStack{} = [t]$, and $\BTStack = []$.
%
The procedure advances by repeatedly moving to a suitable child node:
%
\begin{enumerate}
\item[A.] If the node is labeled with a symbol $\cst{f}$ and the top item
  $t$ of \TermStack{} is \confrep{}{of the form }$\cst{f}(\tuple{t}{n})$,
  replace $t$ by $n$~new items $t_1,\dots,t_n$, and push $t$ onto \BTStack.

\smallskip
\item[B.] If the node is labeled with a variable $x$, there are two subcases.
  If $x$ is already bound, check that $\sigma(x) = t$;
  otherwise, extend $\sigma$ so that $\sigma(x) = t.$ Next, pop the term $t$
  from \TermStack{} and push it onto \BTStack.
\end{enumerate}

%
The goal is to reach an accepting node. If the query term and all the terms
stored in the tree are first-order, \TermStack{} will then be empty, and the
entire query term will have been matched.
%
Backtracking works in reverse: Pop a term $t$ from \BTStack; if the current
node is labeled with an $n$-ary symbol, discard \TermStack{}'s topmost
$n$~items; push $t$ onto \TermStack. Undo any variable bindings.

As an example, looking up $\cst{g}(\cst{b}, \cst{a})$ in the tree $D_1$
would result in the following succession of stack states, starting from the
root $\varepsilon$ along the path $\cst{g}.\cst{b}.\cst{a}$:
%
\[\begin{array}{@{}l@{\kern1.5em}l@{\kern1em}l@{\kern1em}l@{\kern1em}l@{}}
& \varepsilon & \cst{g} & \cst{g}.\cst{b} & \cst{g}.\cst{b}.\cst{a} \\[.5\jot]
\sigma{:} & \emptyset & \emptyset & \emptyset & \emptyset \\
\TermStack{:} & [\cst{g}(\cst{b}, \cst{a})] & [\cst{b}, \cst{a}] & [\cst{a}] & [\,] \\
\BTStack{:} & [\,] & [\cst{g}(\cst{b}{,}\; \cst{a})] & [\cst{b}{,}\; \cst{g}(\cst{b}{,}\; \cst{a})] & [\cst{a}{,}\; \cst{b}{,}\; \cst{g}(\cst{b}, \cst{a})]
\end{array}\]
%
Backtracking amounts to moving leftward:
To get back from $\cst{g}$ to the root, we pop $\cst{g}(\cst{b}, \cst{a})$
from \BTStack, we discard two items from \TermStack{}, and we push
$\cst{g}(\cst{b}, \cst{a})$ onto \TermStack{}.

To adapt the procedure to \lfhol{}, the key idea is that an applied variable
is not very different from an applied symbol. A node labeled with an $n$-ary
head~$\zeta$ matches a prefix~$t'$ of the $k$-ary term $t$
popped from \TermStack{} and leaves $n - k$ arguments $\overline{u}$ to be
pushed back, with $t = t' \; \overline{u}.$ If $\zeta$ is a variable, it must
be bound to the prefix~$t'$ assuming $\zeta$ and $t'$ are of same type.
%
Backtracking works analogously: Given the arity $n$ of the node label $\zeta$
and the arity $k$ of the term $t$ popped from \BTStack{}, we discard the
topmost $n - k$ items $\overline{u}$ from \BTStack.

\looseness=-1
To illustrate the procedure, we consider the tree $D_2$ but change $y$'s type
to $\iota \to \iota.$ This tree stores
$\{
\cst{f}\; x
{,}\allowbreak\;
\cst{g}\; \cst{a}\; \cst{a}
{,}\allowbreak\;
\cst{g}\; (y\; \cst{a})
{,}\allowbreak\;
\cst{g}\; (y\; x)
{,}\allowbreak\;
x
\}.$
Let $\cst{g}\; (\cst{g}\;\cst{a}\;\cst{b})$ be the query term.
We have the following sequence of substitutions $\sigma$ and stacks
$\TermStack, \BTStack$:
%
\[\begin{array}{@{}l@{\kern1em}l@{\kern1em}l@{\kern1em}l@{}}
\varepsilon & \cst{g} & \cst{g}.y & \cst{g}.y.x \\[.5\jot]
\emptyset & \emptyset & \{y \mapsto \cst{g}\> \cst{a}\} & \{y \mapsto \cst{g}\> \cst{a}{,}\; x \mapsto \cst{b}\} \\[0pt]
[\cst{g}\> (\cst{g}\> \cst{a}\> \cst{b})] & [\cst{g}\> \cst{a}\> \cst{b}] & [\cst{b}] & [\,] \\[0pt]
[\,] & [\cst{g}\> (\cst{g}\> \cst{a}\> \cst{b})] & [\cst{g}\> \cst{a}\> \cst{b}{,}\;\cst{g}\> (\cst{g}\> \cst{a}\> \cst{b})] & [\cst{b}{,}\;\cst{g}\> \cst{a}\> \cst{b}{,}\;\cst{g}\> (\cst{g}\> \cst{a}\> \cst{b})]
\end{array}\]

\confrep{}{When backtracking from $\cst{g}.y$ to $\cst{g}$, by comparing
$y$'s arity of $n = 1$ with $\cst{g}\; \cst{a}\; \cst{b}$'s arity of $k = 0$,
we determine that one~item must be discarded from \TermStack{}. }%
Finally, to avoid traversing twice as many subterms as in the first-order
case, we can optimize prefixes: Given a
query term $\zeta\;\tuple{t}{n}$, we can also match prefixes
$\zeta\;\tuple{t}{k}$, where $k < n$, by allowing \TermStack{} to be nonempty when we reach an accepting node.
%If accepting nodes correspond to legal prefixes, the stack will then contain
%exactly the remaining arguments: $[t_{k+1},\dots,t_n].$

\newcommand\quadruple{\mathcalx{Q}}
Similarly to unification and matching, we present finding generalizations
in a perfect discrimination tree as a transition system. States
are quadruples $\quadruple = (\overline{t}, \overline{b}, \mathit{D}, \sigma)$, where
$\overline{t}$ is a list of terms, $\overline{b}$ is a list of tuples storing
backtracking information, $\mathit{D}$ is a discrimination
(sub)tree, and $\sigma$ is a substitution.

%With $+$ we denote list append operation, and with
%$::$ we denote operation of prepending single element to list.

Let $D$ be a perfect discrimination tree. $\Term(D)$ denotes the set of terms
stored in $D$. The function $D|_\zeta$ returns the child of $D$ labeled with
$\zeta$, if it exists. Child nodes are themselves perfect discrimination
(sub)trees. Given any node $D$, if the node is accepting, then the value stored
on that node is defined as $\mathit{val}(D) = (s, d)$, where $s$ is the
accepted term and $d$ is some arbitrary data;
%(typically, an equation whose one side is~$s$)
otherwise, $\mathit{val}(D)$ is undefined.

%We label the tree nodes using BFS, starting with 0 as label for
%root node.

%With $\mathcalx{R}$ we denote $(\overline{t}, \overline{b}, D, \sigma)$ or
%$(\mathit{v}, \sigma)$ or $\bot$.

Starting from an initial state $([t], [\,], D, \emptyset)$, where $t$ is the query
term and $D$ is an entire discrimination tree, the following transitions are
possible:
% \begin{description}[labelwidth=\widthof{\rm\textsf{AdvanceX}}]
%   \item[\rm\textsf{AdvanceF}]
%     $ ( \cst{f} \> \tuple{s}{m} \mathrel{\cdot} \overline{t}{,}\; \overline{b}{,}\; D{,}\; \sigma )
%         \Pdtarrow
%         (\tuple{s}{m} \cdot \overline{t}{,}\;
%         (\cst{f} \> \tuple{s}{m}, D, \sigma) \cdot \overline{b}{,}\; D{|}_\cst{f}{,}\; \sigma)
%       $\quad
%       if $D{|}_\cst{f}$ is defined

%   \smallskip
%   \item[\rm\textsf{AdvanceX}]
%     $ ( s \; \tuple{s}{m} \mathrel{\cdot} \overline{t}{,}\; \overline{b}{,}\; D{,}\; \sigma )
%         \Pdtarrow {}$ \\
%     $(\tuple{s}{m} \cdot \overline{t}{,}\; (s \; \tuple{s}{m}, D, \sigma) \cdot \overline{b}{,}\; D{|}_x{,}\; \sigma[x \mapsto s])
%       $\\
%       if $D{|}_x$ is defined, $x$ and $s$ have the same type, and
%       $\sigma(x)$ is either undefined or equal to $s$

%   \smallskip
%   \item[\rm\textsf{Backtrack}]
%     $ ( \tuple{s}{m} \cdot \overline{t}{,}\; (s, D_0, \sigma_0) \cdot \overline{b}{,}\; D{,}\; \sigma)
%         \Pdtarrow %\\\hphantom{\rm\textsf{AdvanceX}}%
%       (s \mathrel{\cdot} \overline{t}{,}\; \overline{b}{,}\; D_0{,}\; \sigma_0)$ \\
%       if $D_0{|}_\zeta = D$ and $m = \textit{arity}(\zeta) - \textit{arity}(s)$

%   \smallskip
%   \item[\rm\textsf{Success}]
%     $ ( [\,]{,}\; \overline{b}{,}\; D{,}\; \sigma )
%         \Pdtarrow (\mathit{val}(D){,}\; \sigma)$ \\
%     if~$\mathit{val}(D)$ is defined
% \end{description}

\noindent
\begin{tabular}{ll}
  \textsf{AdvanceF} & $ ( \cst{f} \> \tuple{s}{m} \mathrel{\cdot} \overline{t}{,}\; \overline{b}{,}\; D{,}\; \sigma )
           \Pdtarrow
           (\tuple{s}{m} \cdot \overline{t}{,}\;
           (\cst{f} \> \tuple{s}{m}, D, \sigma) \cdot \overline{b}{,}\; D{|}_\cst{f}{,}\; \sigma)
         $ \\
                   & if $D{|}_\cst{f}$ is defined \\[\jot]
  \textsf{AdvanceX} & $( s \; \tuple{s}{m} \mathrel{\cdot} \overline{t}{,}\; \overline{b}{,}\; D{,}\; \sigma ) 
                      \Pdtarrow {} 
                      (\tuple{s}{m} \cdot \overline{t}{,}\; (s \; \tuple{s}{m}, D, \sigma) \cdot \overline{b}{,}\; D{|}_x{,}\; \sigma[x \mapsto s])$ \\
                    & if $D{|}_x$ is defined, $x$ and $s$ have the same type,\\ & and
                    $\sigma(x)$ is either undefined or equal to $s$ \\[\jot]
  \textsf{Backtrack} & $ ( \tuple{s}{m} \cdot \overline{t}{,}\; (s, D_0, \sigma_0) \cdot \overline{b}{,}\; D{,}\; \sigma)
                      \Pdtarrow
                      (s \mathrel{\cdot} \overline{t}{,}\; \overline{b}{,}\; D_0{,}\; \sigma_0)$  \\
                      & if $D_0{|}_\zeta = D$ and $m = \textit{arity}(\zeta) - \textit{arity}(s)$ \\[\jot]
  \textsf{Success}    & $ ( [\,]{,}\; \overline{b}{,}\; D{,}\; \sigma ) \Pdtarrow (\mathit{val}(D){,}\; \sigma)$ \\
                      & if~$\mathit{val}(D)$ is defined
\end{tabular}

Above, $\cdot$ denotes prepending an element or a list to a list.
%
Intuitively, \textsf{AdvanceF} and \textsf{AdvanceX} move deeper in
the tree, generalizing cases A and B above to \lfhol{} terms.
\textsf{Backtrack} can be used to return to a previous state.
\textsf{Success} extracts the term~$t$ and data~$d$ stored in
an accepting node.

\newcommand\BDTARROW[1]{\rlap{\ensuremath{\Pdtarrow_{#1}\;}}\phantom{\Pdtarrow_\textsf{AdvanceX}\;}}

The following derivation illustrates how to locate a generalization of
$\cst{g}\; (\cst{g}\;\cst{a}\;\cst{b})$ in the tree $D_2$:\pagebreak[2]
%
\begin{align*}
 & ([\cst{g} \; (\cst{g} \; \cst{a} \; \cst{b})]{,}\;  [\,]{,}\;  D{,}\;  \emptyset) \\[-1\jot]
 \BDTARROW{\textsf{AdvanceF}}&
 ([\cst{g} \; \cst{a} \; \cst{b}]{,}\;  [(\cst{g} \; (\cst{g} \; \cst{a} \; \cst{b}), D, \emptyset)]{,}\;  D{|}_{\cst{g}}{,}\;  \emptyset)\\[-1\jot]
 \BDTARROW{\textsf{AdvanceX}}&
 ([\cst{b}]{,}\;  [(\cst{g} \; \cst{a} \; \cst{b}, D{|}_{\cst{g}}, \emptyset), \ldots]{,}\;  D{|}_{\cst{g}.y}{,}\;  \{y \mapsto \cst{g} \; \cst{a} \}) \\[-1\jot]
 \BDTARROW{\textsf{AdvanceX}}&
 ([\,]{,}\;  [(\cst{b}, D{|}_{\cst{g}.y}, \{y \mapsto \cst{g} \; \cst{a} \}), \ldots]{,}\; D{|}_{\cst{g}.y.x}{,}\; \{ y \mapsto \cst{g} \; \cst{a}{,}\; x \mapsto \cst{b} \}) \\[-1\jot]
 \BDTARROW{\textsf{Success}}& ((\cst{g} \; (y \; x), d){,}\; \{ y \mapsto \cst{g} \; \cst{a}{,}\; x \mapsto \cst{b} \})
\end{align*}

Let ${\Pdtarrow_{\mathsf{Advance}}} = {\Pdtarrow_{\mathsf{AdvanceF}}} \cup
{\Pdtarrow_{\mathsf{AdvanceX}}}$.
%and ${\Pdtarrow_{\mathsf{NoBacktrack}}} = {\Pdtarrow_{\mathsf{Advance}}} \cup
%{\Pdtarrow_{\mathsf{Success}}}$.
It is easy to show that \textsf{Backtrack} undoes an \textsf{Advance}
transition: % and can be avoided when starting from an initial state.

\begin{lemma}\label{lemma:ehoh:backtrack-inverses-advance} If $\quadruple \Pdtarrow_{\mathsf{Advance}} \quadruple'$, then
  $\quadruple' \Pdtarrow_{\mathsf{Backtrack}} \quadruple$.
  \end{lemma}
\begin{proof}
For both \textsf{Advance} steps, we show that \textsf{Backtrack} restores the state
properly. If \textsf{AdvanceF} was applied, we have
%
\begin{align*}
  ( \cst{f} \> \tuple{s}{m} \mathrel{\cdot} \overline{t}{,}\; \overline{b}{,}\; D{,}\; \sigma )
& \Pdtarrow_{\mathsf{AdvanceF}}
  (\tuple{s}{m} \cdot \overline{t}{,}\;
      (\cst{f} \> \tuple{s}{m}, D, \sigma) \cdot \overline{b}{,}\; D{|}_\cst{f}{,}\; \sigma)
\\[-\jot]
& \Pdtarrow_{\mathsf{Backtrack}}
  ( \overline{t'}{,}\; \overline{b}{,}\; D{,}\; \sigma )
\end{align*}
We must show that $\overline{t'} = \cst{f} \> \tuple{s}{m} \mathrel{\cdot} \overline{t}$.
Let $k = \textit{arity}(\cst{f})$ and $l = \textit{arity}(\cst{f} \> \tuple{s}{m})$.
By definition of $k$, we have $m = k - l$, as in \textsf{Backtrack}'s side
condition. Thus, $\overline{t'} = \cst{f} \> \tuple{s}{m} \mathrel{\cdot}
\overline{t}$. The other case is
\begin{align*}
  ( s \; \tuple{s}{m} \mathrel{\cdot} \overline{t}{,}\; \overline{b}{,}\; D{,}\; \sigma )
& \Pdtarrow_{\mathsf{AdvanceX}}
  (\tuple{s}{m} \cdot \overline{t}{,}\; (s \; \tuple{s}{m}, D, \sigma) \cdot \overline{b}{,}\; D{|}_x{,}\; \sigma')
\\[-\jot]
& \Pdtarrow_{\mathsf{Backtrack}}
  ( \overline{t'}{,}\; \overline{b}{,}\; D{,}\; \sigma )
\end{align*}
where $\sigma' = \sigma[x \mapsto s]$.
Again, we must show that $\overline{t'} = s \; \tuple{s}{m} \mathrel{\cdot} \overline{t}$. Terms
$x$ and $s$ must have the same type for \textsf{AdvanceX} to be applicable; therefore, they have the same arity. Then, we conclude $m=
\textit{arity}(s) - \textit{arity}(s \; \tuple{s}{m} ) =  \textit{arity}(x) - \textit{arity}(s \; \tuple{s}{m} )$, as in \textsf{Backtrack}'s
side condition. Thus, $\overline{t'} = s \; \tuple{s}{m} \mathrel{\cdot} \overline{t}$.
\end{proof}

\begin{lemma}\label{lemma:ehoh:advance-inverses-backtrack}
  If $\quadruple \Pdtarrow_{\mathsf{Advance}} \quadruple'
  \Pdtarrow_{\mathsf{Backtrack}} \quadruple''$, then
  $\quadruple'' = \quadruple$.
  \end{lemma}
\begin{proof}
  By Lemma~\ref{lemma:ehoh:backtrack-inverses-advance}, $\quadruple'
  \Pdtarrow_{\mathsf{Backtrack}} \quadruple$. Furthermore, \textsf{Backtrack} is
  clearly functional. Thus, $\quadruple'' = \quadruple$.
\end{proof}

\begin{lemma}\label{lemma:ehoh:backtrack-remove}
  Let $\quadruple = ([t], [\,], D, \emptyset)$. If $\quadruple
  \Pdtarrow^* \quadruple'$, then $\quadruple \Pdtarrow_{\mathsf{Advance}}^*\allowbreak
  \quadruple'$.
\end{lemma}

\begin{proof}
  \looseness=-1
  Let $\quadruple = \quadruple{}_{\,\,0} \Pdtarrow \cdots \Pdtarrow \quadruple{}_{\,\,n} = \quadruple'$.
  Let $i$ be the index of the first transition of the form $\quadruple{}_{\,\,i} \Pdtarrow_{\mathsf{Backtrack}}
  \quadruple{}_{\,\,i+1}$. Since $\quadruple{}_{\,\,0}$'s backtracking stack is empty, we must
  have $i \not= 0$. Hence, we have $\quadruple{}_{\,\,i-1} \Pdtarrow_{\mathsf{Advance}}
  \quadruple{}_{\,\,i} \Pdtarrow_{\mathsf{Backtrack}} \quadruple{}_{\,\,i+1}$. By
  Lemma~\ref{lemma:ehoh:advance-inverses-backtrack}, $\quadruple{}_{\,\,i-1} =
  \quadruple{}_{\,\,i+1}$. Thus, we can shorten the derivation to $\quadruple{}_{\,\,0}
  \Pdtarrow \cdots \Pdtarrow \quadruple{}_{\,\,i-1} = \quadruple{}_{\,\,i+1} \Pdtarrow
  \cdots \Pdtarrow \quadruple{}_{\,\,n}$,
  thereby eliminating one \textsf{Backtrack} transition. By repeating this
  process, we eliminate all \textsf{Backtrack} transitions.
\end{proof}

\begin{lemma}\label{lemma:ehoh:termination-pdt}
  There exist no infinite chains of the form
  $\quadruple{}_{\,\,0} \Pdtarrow_\mathsf{Advance}\allowbreak \quadruple{}_{\,\,1} \Pdtarrow_\mathsf{Advance}
  \cdots$.
\end{lemma}
\begin{proof}
  With each \textsf{Advance} transition, the height of the discrimination tree
  decreases by at least one.
\end{proof}

Perfect discrimination trees match a single term against a set of terms. To
prove them correct, we will connect them to the transition system
$\Matcharrow$ for matching (Sect.~\ref{sec:ehoh:unif-match}). This connection\pagebreak[2] will
help us show that whenever a discrimination tree stores a generalization of a
query term, this generalization can be found.
%
To express the refinement, we introduce an intermediate transition
system, $\Matchiiarrow$, that focuses on a single pair of terms (like
$\Matcharrow$) but that solves the constraints in a depth-first, left-to-right
fashion and builds the substitution incrementally (like $\Pdtarrow$).
Its initial states are of the form $([s \MATCH t], \emptyset)$. Its transitions
are as follows:

\noindent
\begin{tabular}{ll}
  \textsf{Decompose} & $ (\cst{f} \; \tuple{s}{m} \MATCH \cst{f} \; \tuple{t}{m} \mathrel{\cdot} \overline{c}, \, \sigma) \Matchiiarrow {} ( (s_1 \MATCH t_1, \ldots, s_m \MATCH t_m) \cdot \overline{c}, \sigma ) $ \\[\jot]
  \textsf{DecomposeX} & $(x \; \tuple{s}{{m}} \MATCH u \; \tuple{t}{{m}} \mathrel{\cdot} \overline{c}, \, \sigma) \Matchiiarrow {} ( (s_1 \MATCH t_1, \ldots, s_m \MATCH t_m) \cdot \overline{c}, \sigma[x \mapsto u] )$ \\
                    & if $x$ and $u$ have the same type and either $\sigma(x)$ is undefined or $\sigma(x)=u$ \\[\jot]
  \textsf{Success} & $ ([\,], \, \sigma) \Matchiiarrow  \sigma $ \\[\jot]
  \textsf{Clash}    & $ (\cst{f} \; \tuple{s}{m} \MATCH \cst{g} \; \tuple{t}{n} \mathrel{\cdot} \overline{c}, \, \sigma) \Matchiiarrow \bot $ \\[\jot]
  \textsf{ClashTypeX}    & $(x \; \tuple{s}{{m}} \MATCH u \; \tuple{t}{{m}} \mathrel{\cdot} \overline{c}, \, \sigma) \Matchiiarrow \bot$ \\
                         & if~$x$ and $u$ have different types  \\[\jot]
  \textsf{ClashLenXF}    & $(x \; \tuple{s}{{m}} \MATCH \cst{f} \; \tuple{t}{{n}} \mathrel{\cdot} \overline{c}, \, \sigma) \Matchiiarrow \bot$ \\
                         & if~$m > n$  \\[\jot]
  \textsf{ClashLenXY}    & $(x \; \tuple{s}{{m}} \MATCH y \; \tuple{t}{{n}} \mathrel{\cdot} \overline{c}, \, \sigma) \Matchiiarrow \bot$ \\
                         & if~$x \neq y$ and $m > n$ \\[\jot]
  \textsf{ClashFX}    & $(\cst{f} \; \overline{s} \MATCH x \; \overline{t}  \mathrel{\cdot} \overline{c}, \, \sigma) \Matchiiarrow \bot$ \\[\jot]
  \textsf{Double}    & $(x \; \tuple{s}{{m}} \MATCH u \; \tuple{t}{{m}}  \mathrel{\cdot} \overline{c}, \, \sigma) \Matchiiarrow \bot$ \\
                     & if $x$ and $u$ have the same type, $\sigma(x)$ is defined, and $\sigma(x) \neq u$ \\[\jot]
\end{tabular}

\newcommand\AlphaMatch[1]{\alpha(#1)}

We need an auxiliary function to convert $\Matchiiarrow$ states to
$\Matcharrow$ states.
Let
$\AlphaMatch{\{ x_1 \mapsto s_1,\allowbreak \ldots, x_m \mapsto s_m \}}
 = \{ x_1 \MATCH s_1,\ldots,\,\allowbreak x_m \MATCH s_m \}$,
$\AlphaMatch{\overline{c}, \sigma} =
\{ c \mid c \in \overline{c} \} \mathrel\cup \AlphaMatch{\sigma}$, and
$\AlphaMatch{\bot} = \bot$.
%
Moreover, let $\mathcal{S}$ range over states of the form $(\overline{c}, \sigma)$ and
$\mathcalx{R}$ additionally range over special states of the form $\sigma$ or
$\bot$.

\begin{lemma}\label{lemma:ehoh:match-refinement}
  If $\mathcal{S} \Matchiiarrow \mathcalx{R}$,
  then $\AlphaMatch{\mathcal{S}} \Matcharrow^* \AlphaMatch{\mathcalx{R}}$.
\end{lemma}
\begin{proof}
By case distinction on $\mathcalx{R}$. Let $\mathcal{S} = (\overline{c}, \sigma)$.

\medskip

\begin{sloppypar}%
  \noindent
  \textsc{Case} $\mathcalx{R} = (\overline{c'}, \sigma')$:\enskip
  Only $\Matchiiarrow_{\mathsf{Decompose}}$ and
  $\Matchiiarrow_{\mathsf{DecomposeX}}$ are possible.
  If $\Matchiiarrow_{\mathsf{Decompose}}$ is applied, then $\Matcharrow_{\mathsf{Decompose}}$
  is applicable and results in $\AlphaMatch{\mathcalx{R}}$.
  If $\Matchiiarrow_{\mathsf{DecomposeX}}$ is applied,
  we have either $m>0$, and $\Matcharrow_{\mathsf{DecomposeX}}$ is applicable,
  or $m = 0$, and $\AlphaMatch{\overline{c'},\sigma'} = \AlphaMatch{\mathcal{S}}$,
  which implies that the two states are connected by an idle transition of
  $\Matcharrow^*$.
  \end{sloppypar}

  \medskip

  \noindent
  \textsc{Case} $\mathcalx{R} = \bot$:\enskip
  All the $\Matchiiarrow$ rules resulting in $\bot$ except for \textsf{Double}
  have the same side conditions as the corresponding $\Matcharrow$ rules.
  $\Matchiiarrow_{\mathsf{Double}}$ corresponds to
  $\Matcharrow_{\mathsf{Double}}$ if $m=0$. If $m \neq 0$, we need
  an intermediate $\Matcharrow_{\mathsf{DecomposeX}}$ step before
  $\Matcharrow_{\mathsf{Double}}$ can be applied to derive $\bot$. Since
  $\Matchiiarrow_{\mathsf{Double}}$ is applicable, $\sigma(x) = u'
  \neq u$. Hence, $x \MATCH u'$ must be present in $\AlphaMatch{\overline{c},
  \sigma}$. $\Matcharrow_{\mathsf{DecomposeX}}$ will augment this set with
  $x \MATCH u$, enabling $\Matcharrow_{\mathsf{Double}}$.

  \medskip

  \noindent
  \textsc{Case} $\mathcalx{R} = \sigma$:\enskip
  The only possible rule is $\Matchiiarrow_{\mathsf{Success}}$,
  with $\overline{c}=[\,]$. Since $\AlphaMatch{\mathcal{S}} = \AlphaMatch{\sigma}$,
  this transition corresponds to an idle transition of $\Matcharrow^*$.
\end{proof}

\begin{lemma}\label{lemma:ehoh:refined-match-partial-correctness}\,%
  If $\mathcal{S} \Matchiiarrow^{!} \mathcalx{R}$, then $\mathcalx{R}$ is either
  some substitution $\sigma'$ or $\bot$.
  If $\mathcal{S} \Matchiiarrow^{!} \sigma'$, then $\sigma'$ is the MGG
  of~$\AlphaMatch{\mathcal{S}}$.
  If $\mathcal{S} \Matchiiarrow^{!} \bot$, then $\AlphaMatch{\mathcal{S}}$ has
  no solutions.
  \end{lemma}
\begin{proof}
  %Let $\mathcal{S} = (\overline{c}, \sigma)$.
